Lost in the woods: shifting habitats can lead phylogeography astray

### 2025-03-09 manuscript/phylogeo.tex
change from last year

In phylogeographic inference, one can either assume that individuals migrate between discrete locations or disperse in continuous space.
Here, we consider migration in continuous space, where the underlying model is typically assumed to be diffusive, that is the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
\begin{equation}
    P(\rvec_c| \rvec_p, t) = \frac{1}{4\pi D t}e^{\frac{|r_c - r_p|^2}{4Dt}}
\end{equation}
where $D$ is a diffusivity with dimensions $\mathrm{length}^2/\mathrm{time}$, we have assumed a two-dimensional space, and we have assumed dispersal is isotropic.
Such a diffusive process is the simplest and most natural choice in absence of directed motion or long range migration.
It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree along with the spatial likelihood.
But this model assumes dispersal of one lineage is independent of other lineages, and that dispersal properties are homogeneous across the habitat, and that the habitat does not change over time.
Furthermore, it assumes that the tree topology and the branching rates of the tree are independent of the spatial location.
This independence assumption is for example manifest in coalescent models assuming ``exchangeability'' of individuals, or birth-death models assuming identical rates on contemporary branches of the tree.

@@

Instead of the coalescent trees used above, I now used an individual based simulation forward in time.
Individuals are assigned a fitness $f=\alpha(1-\hat{\phi}/\phi_0) + (1-N/N_0)$ and produce a Poisson distributed number of offspring with mean $1+f$.
The first term determines the fitness due to the local population density and has the tendency to stabilize this density at $\phi_0$, the second term $(1-N/N_0)$ keeps the overall population close to the target population size $N_0$.
The population density is calculated using a Gaussian kernel with interaction radius $r$
\begin{equation}
    \label{eq:density_estimator}
    \hat{\phi}(\rvec, t) = \sum_{i\in N} \frac{e^{-|\rvec_i - \rvec|^2/2r^2}}{\sqrt{2\pi r^2}}
\end{equation}
where the sum is over all individuals $i$ in the population of size $N$ and $\rvec_i$ is location of individual $i$ at time $t$.
The spatial locations of the offspring are chosen from a 2-D Gaussian centered on the parent with a variance to match the diffusion constant.
The simulation records the tree of population, including their ancestral location and times.
In addition to the true location of ancestral nodes, I also infer their locations from the locations of the extant samples.
For this inference, I make the common assumption of continuous phylogeography that growth and location are uncoupled and that dispersal is diffusive.
In this case, the marginal distributions of ancestral locations and their mean and variance can be calculated exactly using recursion relations.

>>>>

In phylogeographic inference, one can either assume that individuals migrate between discrete locations or disperse in continuous space.
Here, we consider migration in continuous space, where the underlying model is typically assumed to be diffusive, that is the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
\begin{equation}
    P(\rvec_c| \rvec_p, t) = \frac{1}{4\pi D t}e^{\frac{|r_c - r_p|^2}{4Dt}}
\end{equation}
where $D$ is a diffusivity with dimensions $\mathrm{length}^2/\mathrm{time}$, we have assumed a two-dimensional space, and we have assumed dispersal is isotropic.
Such a diffusive process is the simplest and most natural choice in absence of directed motion or long range migration.
It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree along with the spatial likelihood.
But this model assumes dispersal of one lineage is independent of other lineages, that dispersal properties are homogeneous across the habitat, and that the habitat does not change over time.
Furthermore, it assumes that the tree topology and the branching rates of the tree are independent of the spatial location.
This independence assumption is for example manifest in coalescent models assuming ``exchangeability'' of individuals, or birth-death models assuming identical rates on contemporary branches of the tree.

@@

Instead of the coalescent trees used above, I now used an individual based simulation forward in time.
Individuals are assigned a fitness $f=\alpha(1-\hat{\phi}/\phi_0) + (1-N/N_0)$ and produce a Poisson distributed number of offspring with mean $1+f$.
The first term determines the fitness due to the local population density and has the tendency to stabilize this density at $\phi_0$, the second term $(1-N/N_0)$ keeps the overall population close to the target population size $N_0$.
The population density is calculated using a Gaussian kernel with interaction radius $r$
\begin{equation}
    \label{eq:density_estimator}
    \hat{\phi}(\rvec, t) = \sum_{i\in N} \frac{e^{-|\rvec_i - \rvec|^2/2r^2}}{\sqrt{4\pi r^2}}
\end{equation}
where the sum is over all individuals $i$ in the population of size $N$ and $\rvec_i$ is location of individual $i$ at time $t$.
The spatial locations of the offspring are chosen from a 2-D Gaussian centered on the parent with a variance to match the diffusion constant.
The simulation records the tree of population, including their ancestral location and times.
In addition to the true location of ancestral nodes, I also infer their locations from the locations of the extant samples.
For this inference, I make the common assumption of continuous phylogeography that growth and location are uncoupled and that dispersal is diffusive.
In this case, the marginal distributions of ancestral locations and their mean and variance can be calculated exactly using recursion relations.
